Generating Correlated Data for Omics Simulation

### 2024-10-18 manuscript/paper.qmd
Edits

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---

@@

Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
We describe and implement three approaches based off decomposing the covariance matrix into a diagonal part and a low-rank part and using a Gaussian copula approach.
These approaches are efficient at omics-scale problems.
We demonstrate the importance of including correlation in two benchmarking applications.
First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics actually improves its performance when dependence is included.
We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.

@@

Recently, we wrote guidelines for performing omics benchmarking [@Brooks2024] and encouraged authors to ensure that any simulated data includes realistic correlation and dependence of measurements within a sample.
However, this highlighted that one reason this is often neglected is that there is relatively little guidance for how to achieve it and that existing solutions are cumbersome.

Here, we give solutions to this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything"), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real dataset and mimicking it.
We implement this strategy as an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
This makes it particularly suited to RNA-seq data but also widely applicable.
Moreover, it supports multi-modal data such as for multi-omics.

@@

We show two applications which demonstrate the importance of including dependence of measurements in simulated data when benchmarking computational pipelines.
First, we simulate RNA-seq data with differential expression between two conditions.
Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported p-values but increased the variance of those estimates.
Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
Surprisingly, we found that including dependence in the data set improved the performance of CYCLOPS.

@@

We chose as a references dataset, 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
For the PCA method, we used $k=2$ and for the spiked Wishart, $k=11$.
The coprcor method always uses the full data matrix, analogous to $k=11$.
Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parameters are not directly comparable across methods.

@@

```{r}
#| label: fig-compare-to-real
#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse cortex data set from GSE151923. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least {HIGH_EXPR_CUTOFF} reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples total for each of dependent and independent simulations) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data.")
#| fig-width: 10
#| fig-height: 10
compare_to_real_plot
```

@@

We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
First, the independent filtering step [@Bourgon2010-uv] assumes independence but has generally be regarded to be robust to typical gene-gene dependence.
Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.

We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, and simulate 5 "experimental" samples.
This was repeated 20 times for each of four dependence conditions (independent, PCA $k=2$, Wishart $k=11$, and corpcor).

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output false discovery rate (FDR) with the true percentages of genes that are differential expressed (@fig-DESeq2 a-b).
We observed that DESeq2 is anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
However, there was a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it performs less consistently on data sets with gene-gene dependence, as in real data sets.

To demonstrate the application of our simulation method for another organism, we also simulated data sets using mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 c-d) as for the fly whole body data sets.

@@

```{r}
#| label: fig-DESeq2
#| fig-cap: "Performance of DESeq2 on simulated datasets.(a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.  Diagonal line represents perfect estimation of FDR. (e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923, Wang et al. 2022), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
#| fig-width: 12
#| fig-height: 6
DESeq2_fdr_plot
```

@@

We further used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures. We use a mouse liver time series data set (GSE151565, Aaronson et al. 2020), which contains a total of 77 samples every 3 hours, for 36 hours. We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point. We then use these to simulate 20 time series data sets with rank $2$ dependence structure, each with 8 time points and 6 samples per time point, and 20 such data sets without dependence.

We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation, defined as follows:
$$\rho = \frac{\sum_{1\leq i<j\leq n}\sin(X_i-X_j) \sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n} \sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n} \sin(Y_i-Y_j)^2)^{1/2}},$$
where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time, respectively, for the $i$-th sample.
$\rho$ has value between $-1$ and $1$, and a $|\rho|$ close to $1$ indicates accurate predictions by CYCLOPS.

We find that CYCLOPS performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is highly impactful on CYCLOPS' performance.
Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig_CYCLOPS b).

@@

Other Gaussian copula-based R packages that may be applicable, at least for datasets with smaller numbers of features, include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods to scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
The `copula` package provides even more flexible dependence options through use of copulas (the NORTA approach is equivalent to using Gaussian copulas).
All of these packages provide more flexibility in specifying dependence than our package, which can only mimic existing datasets, and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---

@@

Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
However, simulation too often ignores these correlations, perhaps due to the inconvenience and computational hurdles of doing so.
We describe and implement three approaches based off decomposing the covariance matrix into a diagonal part and a low-rank part and using a Gaussian copula approach.
These approaches are efficient at omics-scale problems.
We demonstrate the importance of including correlation in two benchmarking applications.
First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
Second, we demonstrate that a method, CYCLOPS, for inferring circadian time of collection from transcriptomics performs substantially worse or better under different dependence situations.
We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.

@@

Recently, we wrote guidelines for performing omics benchmarking [@Brooks2024] and encouraged authors to ensure that any simulated data includes realistic correlation and dependence of measurements within a sample.
However, this highlighted that one reason dependence is often neglected is the relatively lack of guidance for how to achieve it and that existing solutions are cumbersome.

Here, we give solutions to this problem by using a correlation matrix that decomposes into a diagonal part and a low-rank part to both approximate realistic dependencies in a real data set and generate simulated data mimicking the real data.
Using a Gaussian copula approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real dataset and mimicking it.
We implement this strategy as an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.
This makes it particularly suited to RNA-seq data but also widely applicable.
Moreover, it supports multi-modal data such as for multi-omics.

@@

We show two applications which demonstrate the importance of including dependence of measurements in simulated data when benchmarking computational pipelines.
First, we simulate RNA-seq data with differential expression between two conditions.
Using DESeq2 to determine the differentially expressed genes, we found that dependence had little impact on the accuracy of reported $p$-values but increased the variance of those estimates.
Second, we simulated a time series of RNA-seq data points and used the CYCLOPS method to infer collection time from the RNA-seq data, without time labels.
Surprisingly, we found that including dependence in the data set improved the performance of CYCLOPS.

@@

To compare the three simulation methods with a real data set, we chose as a references data set 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
For the PCA method, we used $k=2$ dimensions and for the spiked Wishart, $k=11$.
The coprcor method always uses the full data matrix, analogous to $k=11$.
Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parameters are not directly comparable across methods.

@@

```{r}
#| label: fig-compare-to-real
#| fig-cap: !expr glue::glue("Comparison to real data run on a mouse cortex data set from GSE151923. (a-b) Comparison of gene (a) mean expression and (b) variance, log-scaled in real and PCA simulated data. The line of equality is marked in black. Points are colored according to the density of points in their region. Wishart and corpcor methods give similar results (not shown). (c) Quantile-quantile plot comparing correlation values of gene pairs from real data and simulated data (both with and without dependence). Genes with at least {HIGH_EXPR_CUTOFF} reads were used. Values on the diagonal line indicate a match between the simulated and real data sets. (d) Projections onto the top two principal components of the real data set for both real and simulated data. All 8 simulations (96 samples for each simulation) shown. (e) Principal component analysis was performed on all data sets and the variance captured by the top components is shown. Unlike (d), these components were fit from each data set considered separately instead of reusing the weights from the real data.")
#| fig-width: 10
#| fig-height: 10
compare_to_real_plot
```

@@

We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using data sets simulated with dependence and ones simulated without dependence to compare its performances on both.
DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
First, the independent filtering step [@Bourgon2010-uv] assumes independence but has been reported to be robust to typical gene-gene dependence.
Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.

We used a fly whole body RNA-Seq data set GSE81142 and selected samples of male flies without treatment and after at least 2 hours of feeding to simulate 5 "control" samples.
We then randomly select 5% of the genes to be up or down regulated, with $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down chosen randomly, and simulated 5 "experimental" samples.
This was repeated 20 times for each of four dependence conditions (independent, PCA, Wishart, and corpcor).

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
We observed that DESeq2 is anti-conservative on both data sets, with similar mean true FDRs for each estimated FDR cutoff.
However, there was a greater variance in the performance of DESeq2 on the data sets simulated with dependence, indicating that it performs less consistently on data sets with gene-gene dependence, as in real data sets.

To demonstrate the application of our simulation method for another organism, we also simulated data sets using mouse cortex data set GSE151923 [@Wang2022] and selected samples from male mice.
We then simulated differential expression experiments as above and observed a similar result (@fig-DESeq2 e-f) as for the fly whole body data sets.

@@

```{r}
#| label: fig-DESeq2
#| fig-cap: "Performance of DESeq2 on simulated datasets. (a-d) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the fly whole body data set (GSE81142), (a) without dependence, (b) using PCA, (c) using Wishart and (d) using corpcor.  Diagonal line represents perfect estimation of FDR. (e-h) Comparison of true FDR and DESeq2 reported FDR for data sets simulated from the mouse cortex data set (GSE151923), (e) without dependence, (f) using PCA, (g) using Wishart and (h) using corpcor."
#| fig-width: 12
#| fig-height: 6
DESeq2_fdr_plot
```

@@

We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures. We use a mouse liver time series data set GSE151565, which contains a total of 77 samples every 3 hours, for 36 hours. We compute the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples, and compute the means of gene expressions at each time point.
We then use these to simulate 20 time series data sets for each of the independent, PCA, Wishart, and corpcor simulation methods.

We ran CYCLOPS on each data set with a list of cyclic mouse liver genes (from [@Zhang2014], JTK p-value $<0.05$), which yielded an estimated relative time for each sample.
We evaluated CYCLOPS' performance compared to true circadian time using the circular correlation [@correlation_coeff], defined as follows:
$$\rho = \frac{\sum_{1\leq i<j\leq n}\sin(X_i-X_j) \sin(Y_i-Y_j)}{(\sum_{1\leq i<j\leq n} \sin(X_i-X_j)^2)^{1/2}(\sum_{1\leq i<j\leq n} \sin(Y_i-Y_j)^2)^{1/2}},$$
where $n$ is the number of samples, $X_i$ and $Y_i$ are the true time and CYCLOPS-estimated time, respectively, for the $i$-th sample.
$\rho$ has value between $-1$ and $1$, and a $|\rho|$ close to $1$ indicates accurate predictions by CYCLOPS.

We find that CYCLOPS performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS), indicating that the presence of gene-gene dependence is highly impactful on CYCLOPS' performance.
Overall, CYCLOPS performance is poor in the independent data and improved in both PCA and corpcor data, but surprisingly the Wishart data also shows poor performance.
Since by default CYCLOPS performs dimension reduction to the number of dimensions needed to capture 99% of the variance, this performance difference is likely due to the number of eigengenes needed by CYCLOPS to meet the variance threshold (@fig-CYCLOPS b).

@@

Other Gaussian copula-based R packages that may be applicable, at least for datasets with smaller numbers of features, include `bindata`, `GenOrd`, and `SimMultiCorrData`, the last of these being the most comprehensive.
The `bigsimr` package provides faster implementations of these methods to scale up to omics-level data.
However, even this is computationally demanding; their paper references generating 20,000-dimensional vectors in “under an hour” using 16 threads.
The `copula` package provides even more flexible dependence options through use of copulas.
All of these packages provide more flexibility in specifying dependence than our package, which can only mimic existing datasets, and therefore the longer run-times may be unavoidable for use cases where researchers need to parameterize the dependence structure.

## Data availability

Source code for all simulations and figures in this plot is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
All data used is available from GSE151923, GSE81142, GSE151565.
